---
title: "Generalised influence diagnostics across model types"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Generalised influence diagnostics across model types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

Catch-per-unit-effort (CPUE) standardisation is a cornerstone of fisheries stock
assessment. Two R packages -- `influ` and `influ2` -- have provided influence
diagnostics for GLM- and brms-based CPUE models respectively. The `Ginflu`
package generalises these tools so that the same `get_*` and `plot_*` functions
work across model types, including standard GLMs, Bayesian models fitted with
`brms`, and spatio-temporal models fitted with `sdmTMB`.

This vignette demonstrates the workflow end-to-end:

1. **Simulate** a realistic CPUE data set with zeros and positive catches driven
   by an underlying abundance trend, spatial structure, and operational
   (catchability) covariates. Temporal trends in the vessels present and areas
   fished create strong confounding that the influence diagnostics are designed
   to detect.
2. **Fit GLM-based models** (binomial and lognormal positive) and extract
   indices and influence diagnostics using `Ginflu`.
3. **Fit brms Bayesian models** to the same data and show that `get_index()`,
   `get_step()`, and `plot_bayesian_cdi()` work on `brmsfit` objects.
4. **Fit sdmTMB spatio-temporal models** to the same data and show that the
   same `get_*` / `plot_*` interface applies (CDI plots are not yet supported
   for sdmTMB).

```{r packages}
library(sdmTMB)
library(brms)
library(Ginflu)
library(dplyr)
library(tidyr)
library(ggplot2)
library(cowplot)
library(mvtnorm)
```

# Simulating a CPUE data set

We simulate a 20-year fishery observed across a spatial grid. True abundance
follows a declining-then-recovering trend over time and varies spatially.
Operational covariates (vessel and season) affect catchability. Importantly,
both the probability of capture (encounter rate) and the positive catch rate
depend on abundance -- as abundance declines, encounters become rarer and
positive catches become smaller -- a realistic feature that makes standardisation
essential.

To generate strong, detectable influences, the fleet composition and spatial
coverage shift over time:

- **Vessel trend**: high-catchability vessels (V1--V5) dominate in the early
  years; low-catchability vessels (V11--V15) take over in the later years.
- **Area trend**: productive areas (A1--A2) are fished heavily in the early
  years; poor areas (A5--A6) are fished mainly in the later years.

As a result, the raw CPUE trend is much steeper than the true abundance trend
(inflated early, deflated late), making it an ideal test case for influence
diagnostics.

```{r simulate-data}
set.seed(42)

n_years   <- 20
n_vessels <- 15
n_areas   <- 6
n_seasons <- 4
n_obs     <- 4000

# --- True abundance trend: decline then recovery ---
years <- 1:n_years
true_trend <- exp(-0.04 * (years - 10)^2 / 10 + 0.3 * sin(years / 3))
true_trend <- true_trend / mean(true_trend)  # centre around 1

# --- Structured area effects: A1-A2 productive, A5-A6 poor ---
area_effects <- setNames(
  c(0.55, 0.45, 0.10, -0.10, -0.45, -0.55),
  paste0("A", 1:n_areas)
)

# --- Structured vessel catchability: V1-V5 high, V6-V10 neutral, V11-V15 low ---
vessel_effects <- setNames(
  c(runif(5, 0.5, 0.9), rnorm(5, 0, 0.2), runif(5, -0.9, -0.5)),
  paste0("V", 1:n_vessels)
)

# --- Season effects on catchability ---
season_effects <- setNames(c(-0.3, 0.1, 0.3, -0.1), paste0("S", 1:n_seasons))

# --- Draw observations with year-dependent fleet and area composition ---
# Vessel weights: high-catchability fleet (V1-5) dominant early;
#                 low-catchability fleet (V11-15) dominant late.
# Area weights:   productive areas (A1-2) fished heavily early;
#                 poor areas (A5-6) fished mainly late.
fyear_vec  <- sample(years, n_obs, replace = TRUE)
vessel_col <- character(n_obs)
area_col   <- character(n_obs)

for (yr in years) {
  idx  <- which(fyear_vec == yr)
  n_yr <- length(idx)

  vw <- c(
    rep(pmax(0.05, 1.6 - 0.14 * yr), 5),  # V1-V5:   decreasing over time
    rep(1, 5),                              # V6-V10:  constant
    rep(pmax(0.05, 0.14 * yr - 0.7), 5)   # V11-V15: increasing over time
  )
  vessel_col[idx] <- sample(paste0("V", 1:15), n_yr,
                             prob = vw / sum(vw), replace = TRUE)

  aw <- c(
    rep(pmax(0.05, 1.6 - 0.14 * yr), 2),  # A1-A2: decreasing over time
    rep(1, 2),                             # A3-A4: constant
    rep(pmax(0.05, 0.14 * yr - 0.7), 2)   # A5-A6: increasing over time
  )
  area_col[idx] <- sample(paste0("A", 1:6), n_yr,
                           prob = aw / sum(aw), replace = TRUE)
}

dat <- data.frame(
  fyear  = factor(fyear_vec),
  area   = factor(area_col,   levels = paste0("A", 1:n_areas)),
  vessel = factor(vessel_col, levels = paste0("V", 1:n_vessels)),
  season = factor(sample(paste0("S", 1:n_seasons), n_obs, replace = TRUE))
)

# Local abundance for each observation
dat$log_abundance <- log(true_trend[as.integer(as.character(dat$fyear))]) +
  area_effects[as.character(dat$area)]

# Encounter probability increases with abundance (logit scale)
dat$p_encounter <- plogis(-0.3 + 0.8 * dat$log_abundance +
                            vessel_effects[as.character(dat$vessel)] * 0.3 +
                            season_effects[as.character(dat$season)] * 0.3)

# Positive catch rate also scales with abundance
dat$mu_positive <- exp(2.0 + 0.5 * dat$log_abundance +
                         vessel_effects[as.character(dat$vessel)] +
                         season_effects[as.character(dat$season)])

# Generate catch: zero-inflated lognormal
dat$encounter     <- rbinom(n_obs, 1, dat$p_encounter)
dat$positive_catch <- rlnorm(n_obs, log(dat$mu_positive), 0.6)
dat$catch         <- dat$encounter * dat$positive_catch

# Create positive-only subset and log-catch
dat$log_catch <- ifelse(dat$catch > 0, log(dat$catch), NA)
dat_pos <- dat |> filter(catch > 0)
```

A quick look at the data confirms the structure: many zeros whose proportion
covaries with abundance, and positive catches that track the trend.

```{r data-summary, fig.height = 3.5}
dat |>
  mutate(year = as.integer(as.character(fyear))) |>
  group_by(year) |>
  summarise(
    prop_pos = mean(catch > 0),
    mean_pos = mean(catch[catch > 0]),
    .groups = "drop"
  ) |>
  pivot_longer(-year) |>
  ggplot(aes(year, value)) +
  geom_line() +
  geom_point() +
  facet_wrap(~name, scales = "free_y") +
  labs(x = "Year", y = NULL) +
  theme_minimal()
```

# GLM-based standardisation

## Fitting the models

We fit a binomial GLM for the encounter probability and a lognormal (Gaussian
on log-catch) GLM for positive catches. The first term in each formula is the
year factor (`fyear`), which is the index variable.

```{r glm-fit}
# Binomial model for encounter rate
glm_bin <- glm(encounter ~ fyear + area + vessel + season,
               data = dat, family = binomial(link = "logit"))

# Lognormal model for positive catches
glm_pos <- glm(log_catch ~ fyear + area + vessel + season,
               data = dat_pos, family = gaussian(link = "identity"))
```

## Extracting indices with `get_index`

`get_index()` works on any supported model class. It returns a data frame with
unstandardised and standardised index values plus confidence intervals.

```{r glm-index}
idx_bin <- get_index(glm_bin)
idx_pos <- get_index(glm_pos)
```

```{r glm-index-table}
head(idx_bin)
head(idx_pos)
```

## Index plots with `plot_index`

`plot_index()` takes the output of `get_index()` and produces a standardised
index plot with the unstandardised series overlaid.

```{r glm-plot-index-bin, fig.cap = "Binomial (encounter) index from GLM."}
plot_index(idx_bin)
```

```{r glm-plot-index-pos, fig.cap = "Positive catch index from GLM."}
plot_index(idx_pos)
```

## Step-plots with `get_step` and `plot_step2`

The step-plot shows how the standardised index changes as covariates are
sequentially added to the model. `get_step()` refits the model with terms added
one at a time and collects the index at each step.

```{r glm-step-bin}
step_bin <- get_step(glm_bin)
```

```{r glm-step-bin-plot, fig.height = 8, fig.cap = "Step-plot for the binomial GLM."}
plot_step2(step_bin)
```

```{r glm-step-pos}
step_pos <- get_step(glm_pos)
```

```{r glm-step-pos-plot, fig.height = 8, fig.cap = "Step-plot for the positive GLM."}
plot_step2(step_pos)
```

## CDI plots with `plot_cdi`

The CDI (Coefficient-Distribution-Influence) plot provides a diagnostic summary
for each covariate: estimated coefficients (top-left), data distribution
across years (bottom-left), and the year-specific influence of the covariate
on the index (bottom-right).

```{r glm-cdi-bin, fig.height = 7, fig.width = 8, fig.cap = "CDI plots for the binomial GLM.", results = "hide"}
plot_cdi(glm_bin)
```

```{r glm-cdi-pos, fig.height = 7, fig.width = 8, fig.cap = "CDI plots for the positive GLM.", results = "hide"}
plot_cdi(glm_pos)
```

# brms Bayesian standardisation

`Ginflu` has first-class support for Bayesian models fitted with `brms`. The
same `get_index()` / `get_step()` / `plot_index()` / `plot_step2()` functions
accept `brmsfit` objects directly. For CDI diagnostics, the Bayesian-specific
`plot_bayesian_cdi()` function leverages the full posterior distribution.

## Fitting the models

We fit the same model structure as the GLM section. Keeping chains and
iterations low makes the vignette practical to knit; for real analyses use at
least 4 chains and 2000 iterations.

```{r brms-fit, cache = TRUE, results = "hide"}
brm_bin <- brm(
  encounter ~ fyear + area + vessel + season,
  data    = dat,
  family  = bernoulli(link = "logit"),
  chains  = 2, iter = 1000, warmup = 500,
  cores   = 2, seed = 42, refresh = 0
)

brm_pos <- brm(
  log_catch ~ fyear + area + vessel + season,
  data    = dat_pos,
  family  = gaussian(link = "identity"),
  chains  = 2, iter = 1000, warmup = 500,
  cores   = 2, seed = 42, refresh = 0
)
```

## Extracting indices with `get_index`

`get_index()` draws from the posterior predictive distribution and returns the
same tidy data frame as for GLMs.

```{r brms-index, cache = TRUE}
idx_brm_bin <- get_index(brm_bin)
idx_brm_pos <- get_index(brm_pos)
```

```{r brms-index-table}
head(idx_brm_bin)
head(idx_brm_pos)
```

## Index plots with `plot_index`

```{r brms-plot-index-bin, fig.cap = "Binomial (encounter) index from brms."}
plot_index(idx_brm_bin)
```

```{r brms-plot-index-pos, fig.cap = "Positive catch index from brms."}
plot_index(idx_brm_pos)
```

## Step-plots with `get_step` and `plot_step2`

`get_step()` refits the model with terms added sequentially. Because each
update requires a new MCMC run, this step is slower for brms than for GLMs;
caching is recommended.

```{r brms-step-bin, cache = TRUE}
step_brm_bin <- get_step(brm_bin)
```

```{r brms-step-bin-plot, fig.height = 8, fig.cap = "Step-plot for the brms binomial model."}
plot_step2(step_brm_bin)
```

```{r brms-step-pos, cache = TRUE}
step_brm_pos <- get_step(brm_pos)
```

```{r brms-step-pos-plot, fig.height = 8, fig.cap = "Step-plot for the brms positive model."}
plot_step2(step_brm_pos)
```

## Bayesian CDI plots with `plot_bayesian_cdi`

`plot_bayesian_cdi()` is the Bayesian counterpart of `plot_cdi()`. It uses the
posterior samples of the coefficients and the influence metric to draw violin
plots for categorical covariates (or ribbon plots for continuous ones), together
with the data distribution bubble chart.

```{r brms-cdi-vessel-bin, fig.height = 7, fig.width = 8, fig.cap = "Bayesian CDI for vessel effect (binomial model)."}
plot_bayesian_cdi(brm_bin, xfocus = "vessel", yfocus = "fyear")
```

```{r brms-cdi-area-bin, fig.height = 7, fig.width = 8, fig.cap = "Bayesian CDI for area effect (binomial model)."}
plot_bayesian_cdi(brm_bin, xfocus = "area", yfocus = "fyear")
```

```{r brms-cdi-vessel-pos, fig.height = 7, fig.width = 8, fig.cap = "Bayesian CDI for vessel effect (positive model)."}
plot_bayesian_cdi(brm_pos, xfocus = "vessel", yfocus = "fyear")
```

# sdmTMB spatio-temporal standardisation

The same `get_*` and `plot_*` functions extend to spatio-temporal models fitted
with `sdmTMB`. We add spatial coordinates to our simulated data and fit a
delta (hurdle) model.

## Preparing spatial data

```{r sdm-spatial-setup}
# Assign coordinates to areas (a simple regular layout)
area_coords <- data.frame(
  area = paste0("A", 1:n_areas),
  X = rep(1:3, 2),
  Y = rep(1:2, each = 3)
)

dat <- dat |> left_join(area_coords, by = "area")
dat_pos <- dat |> filter(catch > 0)

# Create the SPDE mesh
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 0.5)

# Prediction grid: all area centroids x all years
pred_grid <- expand.grid(
  X = unique(area_coords$X),
  Y = unique(area_coords$Y),
  fyear = factor(1:n_years)
) |>
  left_join(
    dat |>
      group_by(area) |>
      summarise(across(c(vessel, season), ~ names(sort(-table(.x)))[1]),
                .groups = "drop") |>
      left_join(area_coords, by = "area"),
    by = c("X", "Y")
  )
pred_grid$vessel <- factor(pred_grid$vessel, levels = levels(dat$vessel))
pred_grid$season <- factor(pred_grid$season, levels = levels(dat$season))
```

## Fitting the delta model

We fit a delta (hurdle) sdmTMB model with the same fixed-effect structure.
The first component models encounter probability (binomial), the second models
positive catch rates (lognormal). Spatial random fields account for
unmeasured spatial variation.

```{r sdm-fit, results = "hide"}
fit_sdm <- sdmTMB(
  catch ~ fyear + vessel + season,
  data = dat,
  mesh = mesh,
  family = delta_lognormal(),
  spatial = "on",
  spatiotemporal = "off",
  time = "fyear"
)
```

```{r sdm-sanity}
sanity(fit_sdm)
```

## Extracting indices

`get_index()` recognises the sdmTMB model class. The `predictor` argument
selects the model component: `1` for binomial, `2` for positive.

```{r sdm-index}
idx_sdm_bin <- get_index(fit_sdm, pred_grid = pred_grid, predictor = 1)
idx_sdm_pos <- get_index(fit_sdm, pred_grid = pred_grid, predictor = 2)
```

## Index plots

The same `plot_index()` function works seamlessly.

```{r sdm-plot-index-bin, fig.cap = "Binomial index from sdmTMB."}
plot_index(idx_sdm_bin)
```

```{r sdm-plot-index-pos, fig.cap = "Positive index from sdmTMB."}
plot_index(idx_sdm_pos)
```

## Step-plots

`get_step()` also handles sdmTMB models. It sequentially adds terms and,
for sdmTMB, also adds the spatial and spatiotemporal fields as final steps.

```{r sdm-step-bin}
step_sdm_bin <- get_step(fit_sdm, pred_grid = pred_grid, predictor = 1)
```

```{r sdm-step-bin-plot, fig.height = 8, fig.cap = "Step-plot for the sdmTMB binomial component."}
plot_step2(step_sdm_bin)
```

```{r sdm-step-pos}
step_sdm_pos <- get_step(fit_sdm, pred_grid = pred_grid, predictor = 2)
```

```{r sdm-step-pos-plot, fig.height = 8, fig.cap = "Step-plot for the sdmTMB positive component."}
plot_step2(step_sdm_pos)
```

> **Note:** CDI plots are not yet implemented for sdmTMB models. This
> functionality will be added in a future release.

# Summary

The `Ginflu` package provides a unified interface for CPUE index extraction and
influence diagnostics across model frameworks:

| Function | GLM | brms | sdmTMB | Description |
|:---|:---:|:---:|:---:|:---|
| `get_index()` | Y | Y | Y | Standardised and unstandardised indices |
| `get_step()` | Y | Y | Y | Indices with terms added sequentially |
| `get_influ_glm()` | Y | | | Influence metric for GLM-type models |
| `plot_index()` | Y | Y | Y | Index time-series with confidence bands |
| `plot_step2()` | Y | Y | Y | Step-plot of sequential model building |
| `plot_cdi()` | Y | | | Coefficient-Distribution-Influence diagnostic |
| `plot_bayesian_cdi()` | | Y | | Bayesian CDI with posterior distributions |

By keeping the function signatures consistent, analysts can switch between a
simple GLM-based workflow, a Bayesian workflow, and a full spatio-temporal
standardisation without changing their diagnostic pipeline.
